Statistical analysis script

### 2023-04-05 code/analysis_code/exploratory_analysis.qmd
update linear&Poisson regression model and add random forest

## Save Graph
image_path <- here("results")
ggsave("time_v_hospitalization_pathogen_type.png", path = image_path, width = 3000, height = 2000, units = "px")

>>>>

## Save Graph
image_path <- here("results","images_EDA")
ggsave("time_v_hospitalization_pathogen_type.png", path = image_path, width = 3000, height = 2000, units = "px")

### 2023-04-05 code/analysis_code/statistical_analysis.qmd
update according to comments
update linear&Poisson regression model and add random forest

# Load libraries
```{r}
# Pathing
library(here)

# Data Handling and Modeling
library(tidyverse)
library(tidymodels)
library(poissonreg)
```
# Load Data
```{r}
# Path to Data
data_location <- here(
  "data","processed_data","processed_training_data.rds")
# Import Data
training_data_final <- readRDS(data_location)
glimpse(training_data_final)
```
# First Health Outcome of Interest: Illnesses

@@

```{r}
# Checking Illness Distribution
training_data_final %>%
  ggplot(aes(Illnesses))+
  geom_histogram()+
  scale_x_continuous(trans=scales::pseudo_log_trans(base=10))
image_path <- here("results")
ggsave("histogram_Illnesses.png", path = image_path, width = 3000, height = 2000, units = "px")

# Tibble with Predictors of Interest
illness_data=training_data_final %>%
  select(c(1, 4, 5, 7, 10, 13, 14)) %>%
  mutate_if(is.character, as.factor)
glimpse(illness_data)
```

@@

```{r}
# Linear Regression
lm_spec = linear_reg() %>%
  set_engine(engine = "lm")
## Modeling with Main Predictor `Simplified_Etiology`
lm_fit1 = lm_spec %>%
  fit(Illnesses ~ Simplified_Etiology, data = illness_data)
## Modeling with All Predictors
lm_fit2 = lm_spec %>%
  fit(Illnesses ~ ., data = illness_data)

# Poisson Regression
## Modeling with Main Predictor `Simplified_Etiology`
poisson_wf1 = workflow(Illnesses ~ Simplified_Etiology, poisson_reg())
## Modeling with All Predictors
poisson_wf2 = workflow(Illnesses ~ ., poisson_reg())
## Model Fitting
poisson_fit1 = fit(poisson_wf1, data=illness_data)
poisson_fit2 = fit(poisson_wf2, data=illness_data)
```

### Model Evaluation
```{r}
# Checking Parameters for Linear Regression Model
tidy(lm_fit1) %>%
  arrange(-estimate)
tidy(lm_fit2) %>%
  arrange(-estimate)

# Checking Parameters for Poisson Regression Model
tidy(poisson_fit1) %>%
 arrange(-estimate)
tidy(poisson_fit2) %>%
 arrange(-estimate)

# Model Evaluation
Result = lm_fit1 %>% 
  predict(new_data = illness_data) %>% 
  mutate(.pred = round(.pred), 
        true_value = illness_data$Illnesses, model= "lm_single") %>% 
  bind_rows(lm_fit1 %>% 
              predict(new_data=illness_data) %>% 
              mutate(.pred=round(.pred), 
                     true_value = illness_data$Illnesses, model= "lm_multiple") %>% 
              bind_rows(poisson_fit1 %>% predict(new_data= illness_data) %>%
                          mutate(true_value=illness_data$Illnesses, 
                                 model= "glm_single") %>% 
                          bind_rows(poisson_fit2 %>% 
                                      predict(new_data= illness_data)%>% 
                                      mutate(true_value=illness_data$Illnesses, 
                                             model= "glm_multiple")))
)

Result <- Result %>%
  group_by(model) %>%
  rmse(truth=true_value, estimate = .pred) %>%
  rename(illness_model = model)

saveRDS(Result, file = here("results", "illness_model_evaluation.rds"))

```
According to the result, the model built using Poisson regression with multiple predictors seems to have the smallest root mean squared error.

@@

```{r}
# Checking the Distribution of Hospitalization
training_data_final %>%
  ggplot(aes(Hospitalizations)) +
  geom_histogram() +
  scale_x_continuous(trans=scales::pseudo_log_trans(base = 10))
# Tibble for Variables of Interest
hosp_data = training_data_final %>%
  select(c(1, 4, 5, 7, 11, 13, 14))%>%
  mutate_if(is.character, as.factor)
glimpse(hosp_data)
ggsave("histogram_Hospitalizations.png", path = image_path)
```

@@

```{r}
# Model Specification and Fitting for Linear Regression Model
lm_spec = linear_reg() %>%
  set_engine(engine = "lm")
## Modeling with Main Predictor `Simplified_Etiology`
lm_fit3 = lm_spec %>%
  fit(Hospitalizations ~ Simplified_Etiology, data = hosp_data)
## Modeling with All Predictors
lm_fit4 = lm_spec %>%
  fit(Hospitalizations ~ ., data = hosp_data)

# Model Specification and Fitting for Poisson Regression Model
## Modeling with Main Predictor `Simplified_Etiology`
poisson_wf3 = workflow(Hospitalizations ~ Simplified_Etiology, poisson_reg())
## Modeling with all predictors
poisson_wf4=workflow(Hospitalizations~., poisson_reg())
## Fitting models
poisson_fit3 = fit(poisson_wf3,data = hosp_data)
poisson_fit4 = fit(poisson_wf4,data = hosp_data)
```

### Model Evaluation
```{r}
Result = lm_fit3 %>%
 predict(new_data = hosp_data) %>%
 mutate(.pred = round(.pred), true_value = hosp_data$Hospitalizations,
        model= "lm_single") %>% 
  bind_rows(lm_fit4 %>% 
              predict(new_data=hosp_data) %>% 
              mutate (.pred=round(.pred), 
                      true_value=hosp_data$Hospitalizations, 
                      model= "lm_multiple") %>% 
              bind_rows(poisson_fit1 %>% 
                        predict(new_data = hosp_data) %>%
                        mutate(true_value = hosp_data$Hospitalizations, 
                               model= "glm_single") %>% 
                          bind_rows(poisson_fit2 %>% 
                                      predict(new_data= hosp_data) %>% 
                                      mutate(true_value = hosp_data$Hospitalizations, 
                                             model= "glm_multiple")))
)

Result <- Result %>%
  group_by(model)%>%
  rmse(truth = true_value, estimate = .pred) %>%
  rename(hospital_model = model)

saveRDS(Result, file = here("results", "hospital_model_evaluation.rds"))

@@

According to the result, the model built using linear regression with multiple predictors seems to have the smallest root mean squared error.

>>>>

---
title: "Statistical analysis script"
author: "Weifan Wu and Kailin Chen"
date: "03/16/2023"
output: html_document
---

## Load libraries
```{r}
# Pathing
library(here)

# Data Handling and Modeling
library(tidyverse)
library(tidymodels)
library(poissonreg)
library(ranger)
library(knitr)
library(skimr)
```
## Load Data
```{r}
# Path to Data
data_location <- here(
  "data","processed_data","processed_training_data.rds")
# Import Data
training_data_final <- readRDS(data_location)
skim(training_data_final)
```
## First Health Outcome of Interest: Illnesses

@@

```{r}
# Checking Illness Distribution
training_data_final %>%
  ggplot(aes(Illnesses))+
  geom_histogram()+
  scale_x_continuous(trans=scales::pseudo_log_trans(base=10))
images_stats_path=here("results","images_statistic_analysis")
ggsave("histogram_Illnesses.png", path = images_stats_path, width = 3000, height = 2000, units = "px")

# Create a tibble with predictors of interest
illness_data=training_data_final %>%
  select(c(1, 4, 5, 7, 10, 14)) %>%
  filter(!`IFSAC Category`%in%c("Multiple","Other"),Location!="Unknown")%>%
  mutate_if(is.character, as.factor)
skim(illness_data)
```

@@

```{r}
set.seed(123)
# Set up cross validation
illness_fold=vfold_cv(illness_data,repeats = 2,strata = Illnesses)

# Data pre-processing
## Recipe for main predictor
illness_main_rec=recipe(Illnesses~Simplified_Etiology,data=illness_data)%>%
  step_dummy(Simplified_Etiology)
## Recipe for all predictors
### Lump small proportion into "other" using step_other
### Create dummy code for nominal variables
illness_all_rec=recipe(Illnesses~.,data=illness_data)%>%
  step_other(State,Location,`IFSAC Category`,Simplified_Etiology,threshold = 0.01)%>%
  step_dummy(all_nominal())
### Checking processed data set
illness_all_rec%>%
  prep()%>%
  bake(new_data=NULL)

# Model specification
## Linear Regression
lm_spec = linear_reg() %>%
  set_engine(engine = "lm")

# Creating workflow
## Main Predictor `Simplified_Etiology`
lm_wf1 = workflow() %>%
 add_recipe(illness_main_rec)%>%
  add_model(lm_spec)
## Modeling with All Predictors
lm_wf2 = workflow() %>%
  add_recipe(illness_all_rec)%>%
  add_model(lm_spec)
## Poisson Regression
## Modeling with Main Predictor `Simplified_Etiology`
poisson_wf1 = workflow()%>%
  add_recipe(illness_main_rec)%>%
  add_model(poisson_reg())
## Modeling with All Predictors
poisson_wf2 = workflow()%>%
  add_recipe(illness_all_rec)%>%
  add_model(poisson_reg())
```

## Model fitting
```{r}
## Linear regression
### Main predictor
set.seed(234)
lm_rs1=lm_wf1%>%
fit_resamples(resamples=illness_fold,
    control=control_resamples(save_pred=TRUE))

### All predictors
lm_rs2=lm_wf2%>%
  fit_resamples(resamples=illness_fold,
  control=control_resamples(save_pred=TRUE))

## Poisson regression
### Main predictor
poisson_rs1 = poisson_wf1%>%
  fit_resamples(resamples=illness_fold,
    control=control_resamples(save_pred=TRUE))

### All predictors
poisson_rs2 = poisson_wf2%>%
  fit_resamples(resamples=illness_fold,
    control=control_resamples(save_pred=TRUE))
```

### Model Evaluation
```{r}
# Checking metrics for Linear Regression Model
lm_rs1%>%
  collect_metrics()
lm_rs2%>%
  collect_metrics()
# Checking metrics for poisson regression model
poisson_rs1%>%
  collect_metrics()
poisson_rs2%>%
  collect_metrics()
# Checking Parameters for Linear Regression Model
lm_fit1=lm_wf1%>%
  fit(illness_data)%>%
  tidy()%>%
  arrange(-estimate)

lm_fit2=lm_wf2%>%
  fit(illness_data)%>%
  tidy()%>%
  arrange(-estimate)

# Checking Parameters for Poisson Regression Model
poisson_fit1=poisson_wf1%>%
  fit(illness_data)%>%
  tidy()%>%
  arrange(-estimate)

poisson_fit2=poisson_wf2%>%
  fit(illness_data)%>%
  tidy()%>%
  arrange(-estimate)

# Plotting predictions vs Illnesses for Linear model
lm_rs1%>%
  collect_predictions()%>%
  filter(Illnesses<300)%>%
  ggplot(aes(Illnesses,.pred,color=id2,group=id2))+
  geom_point()+
  geom_smooth(se=FALSE)+
  labs(y="Predicted Illnesses",title="Predicted Illnesses VS Illnesses by linear regression with main predictor")
ggsave("Predicted Illnesses VS Illnesses by linear regression with main predictor.png",path=images_stats_path,width=3000,height=2000,units="px")
lm_rs2%>%
  collect_predictions()%>%
  #filter(Illnesses<300)%>%
  ggplot(aes(Illnesses,.pred,color=id2,group=id2))+
  geom_point()+
  geom_smooth(se=FALSE)+
  labs(y="Predicted Illnesses",title="Predicted Illnesses VS Illnesses by linear regression with all predictors")
ggsave("Predicted Illnesses VS Illnesses by linear regression with all predictors.png",path=images_stats_path,width=3000,height=2000,units="px")
# Plotting predictions vs Illnesses for Poisson model
poisson_rs1%>%
  collect_predictions()%>%
  #filter(Illnesses<300)%>%
  ggplot(aes(Illnesses,.pred,color=id2,group=id2))+
  geom_point()+
  geom_smooth(se=FALSE)+
  labs(y="Predicted Illnesses",title="Predicted Illnesses VS Illnesses by linear regression with main predictor")
ggsave("Predicted Illnesses VS Illnesses by poisson regression with main predictor.png",path=images_stats_path,width=3000,height=2000,units="px")
poisson_rs2%>%
  collect_predictions()%>%
  #filter(Illnesses<500)%>%
  ggplot(aes(Illnesses,.pred,color=id2,group=id2))+
  geom_point()+
  geom_smooth(se=FALSE)+
  labs(y="Predicted Illnesses",title="Predicted Illnesses VS Illnesses by linear regression with all predictors")
ggsave("Predicted Illnesses VS Illnesses by poisson regression with all predictors.png",path=images_stats_path,width=3000,height=2000,units="px")

```
According to the result, the model built using Poisson regression with multiple predictors seems to have the smallest root mean squared error and highest R^2. However, none of those models can catch the outliers (Illness>1000)

@@

```{r}
# Checking the Distribution of Hospitalization
training_data_final %>%
  ggplot(aes(Hospitalizations)) +
  geom_histogram() +
  scale_x_continuous(trans=scales::pseudo_log_trans(base = 10))
# Tibble for Variables of Interest
hosp_data = training_data_final %>%
  select(c(1, 4, 5, 7, 11, 14))%>%
  filter(!`IFSAC Category`%in%c("Multiple","other"),!Location%in%c("Unknown"))%>%
  mutate_if(is.character, as.factor)
skim(hosp_data)
ggsave("histogram_Hospitalizations.png", path = images_stats_path)
```

@@

```{r}
set.seed(1234)
# Set up cross validation
hosp_fold=vfold_cv(hosp_data,repeats = 2,strata = Hospitalizations)

# Data pre-processing
## Recipe for main predictor
hosp_main_rec=recipe(Hospitalizations ~ Simplified_Etiology,data=hosp_data)%>%
  step_dummy(Simplified_Etiology)
## Recipe for all predictors
### Lump small proportion into "other" using step_other
### Create dummy code for nominal variables
hosp_all_rec=recipe(Hospitalizations ~.,data=hosp_data)%>%
  step_other(State,Location,`IFSAC Category`, Simplified_Etiology,threshold = 0.01)%>%
  step_dummy(all_nominal_predictors())
### Checking processed data set
hosp_all_rec%>%
  prep()%>%
  bake(new_data=NULL)

# Model specification
## Linear Regression
lm_spec = linear_reg() %>%
  set_engine(engine = "lm")

# Creating workflow
## Main Predictor `Simplified_Etiology`
lm_wf3 = workflow() %>%
 add_recipe(hosp_main_rec)%>%
  add_model(lm_spec)
## Modeling with All Predictors
lm_wf4 = workflow() %>%
  add_recipe(hosp_all_rec)%>%
  add_model(lm_spec)
## Poisson Regression
## Modeling with Main Predictor `Simplified_Etiology`
poisson_wf3 = workflow()%>%
  add_recipe(hosp_main_rec)%>%
  add_model(poisson_reg())
## Modeling with All Predictors
poisson_wf4 = workflow()%>%
  add_recipe(hosp_all_rec)%>%
  add_model(poisson_reg())
```
## Model fitting
```{r}
## Linear regression
### Main predictor
set.seed(2345)
lm_rs3=lm_wf3%>%
fit_resamples(resamples=hosp_fold,
    control=control_resamples(save_pred=TRUE))

### All predictors
lm_rs4=lm_wf4%>%
  fit_resamples(resamples=hosp_fold,
  control=control_resamples(save_pred=TRUE))

## Poisson regression
### Main predictor
poisson_rs3 = poisson_wf3%>%
  fit_resamples(resamples=hosp_fold,
    control=control_resamples(save_pred=TRUE))

### All predictors
poisson_rs4 = poisson_wf4%>%
  fit_resamples(resamples=hosp_fold,
    control=control_resamples(save_pred=TRUE))
```
### Model Evaluation
```{r}
# Checking metrics for Linear Regression Model
lm_rs3%>%
  collect_metrics()
lm_rs4%>%
  collect_metrics()
# Checking metrics for poisson regression model
poisson_rs3%>%
  collect_metrics()
poisson_rs4%>%
  collect_metrics()
# Checking Parameters for Linear Regression Model
lm_fit3=lm_wf3%>%
  fit(hosp_data)%>%
  tidy()%>%
  arrange(-estimate)

lm_fit4=lm_wf4%>%
  fit(hosp_data)%>%
  tidy()%>%
  arrange(-estimate)

# Checking Parameters for Poisson Regression Model
poisson_fit3=poisson_wf3%>%
  fit(hosp_data)%>%
  tidy()%>%
  arrange(-estimate)

poisson_fit4=poisson_wf4%>%
  fit(hosp_data)%>%
  tidy()%>%
  arrange(-estimate)

# Plotting predictions vs Hospitalizations for Linear model
lm_rs3%>%
  collect_predictions()%>%
  #filter(Hospitalizations<50)%>%
  ggplot(aes(Hospitalizations,.pred,color=id2))+
  geom_point()+
  geom_smooth(se=FALSE)+
  labs(y="Predicted Hospitalizations",title="Predicted Hospitalizations VS Hospitalizations \nby linear regression with main predictor",color="Fold")
ggsave("Predicted Hospitalizations VS Hospitalizations by linear regression with main predictor.png",path=images_stats_path,width=3000,height=2000,units="px")
lm_rs4%>%
  collect_predictions()%>%
  #filter(Hospitalizations<50)%>%
  ggplot(aes(Hospitalizations,.pred,color=id2))+
  geom_point()+
  geom_smooth(se=FALSE)+
  labs(y="Predicted Hospitalizations",title="Predicted Hospitalizations VS Hospitalizations \nby linear regression with all predictors",color="Fold")
ggsave("Predicted Hospitalizations VS Hospitalizations by linear regression with all predictors.png",path=images_stats_path,width=3000,height=2000,units="px")
# Plotting predictions vs Hospitalizations for Poisson model
poisson_rs3%>%
  collect_predictions()%>%
  #filter(Hospitalizations<50)%>%
  ggplot(aes(Hospitalizations,.pred,color=id2))+
  geom_point()+
  geom_smooth(se=FALSE)+
  labs(y="Predicted Hospitalizations",title="Predicted Hospitalizations VS Hospitalizations \nby linear regression with main predictor",color="Fold")
ggsave("Predicted Hospitalizations VS Hospitalizations by poisson regression with main predictor.png",path=images_stats_path,width=3000,height=2000,units="px")
poisson_rs4%>%
  collect_predictions()%>%
  #filter(Hospitalizations<50)%>%
  ggplot(aes(Hospitalizations,.pred,color=id2))+
  geom_point()+
  geom_smooth(se=FALSE)+
  labs(y="Predicted Hospitalizations",title="Predicted Hospitalizations VS Hospitalizations \nby linear regression with all predictors",color="Fold")
ggsave("Predicted Hospitalizations VS Hospitalizations by poisson regression with all predictors.png",path=images_stats_path,width=3000,height=2000,units="px")

@@

According to the result, the model built using Poisson regression with multiple predictors seems to have the smallest root mean squared error and highest R^2 (34%). But again, outliers (Hospitalizations > 100) can not be predicted well by none of those models

## Machine learning approach
## Random Forest
### Predicting First Outcome `Illnesses` using all predictors
```{r}
# Model Specification
rf_spec=rand_forest(mtry=tune(),min_n=tune(),trees=1000)%>%
  set_engine("ranger")%>%
  set_mode("regression")
illness_all_rec=recipe(Illnesses~.,data=illness_data)%>%
  step_other(State,Location,`IFSAC Category`,Simplified_Etiology,threshold = 0.01)%>%
  step_dummy(all_nominal())
# Creating workflow
rf_wf1=workflow()%>%
  add_recipe(illness_all_rec)%>%
  add_model(rf_spec)
# Training and tuning model using space filling design
set.seed(345)
doParallel::registerDoParallel()
rf_tune1=rf_wf1%>%
  tune_grid(
    resamples=illness_fold,
    grid=15,
    control=control_grid(save_pred=TRUE),
    metrics = metric_set(rmse)
  )
```
### Model evaluation
```{r}
rf_tune1%>%
  autoplot()
rf_best1=rf_tune1%>%
  select_best()
rf_final1=rf_wf1%>%
  finalize_workflow(rf_best1)
rf_fit1=rf_final1%>%
  fit(illness_data)
rf_predict1=rf_fit1%>%
  predict(illness_data)
illness_data%>%
  select(Illnesses)%>%
  bind_cols(rf_predict1)%>%
  ggplot(aes(Illnesses,.pred))+
  geom_point()+
  geom_abline(intercept = 0,slope=1)+
  geom_smooth(se=FALSE)+
  labs(y="Predicted Illnesses",title="Predicted illnesses VS illnesses by random forest model")
ggsave("Predicted illnesses VS illnesses by random forest model.png",path=images_stats_path,width=3000,height=2000,units="px")
```

### Predicting second outcome `Hospitalizations` using all predictors
```{r}

# Creating workflow
rf_wf2=workflow()%>%
  add_recipe(hosp_all_rec)%>%
  add_model(rf_spec)
# Training and tuning model
doParallel::registerDoParallel()
set.seed(456)
rf_tune2=rf_wf2%>%
  tune_grid(
    resamples=hosp_fold,
    grid=15,
    control=control_grid(save_pred=TRUE),
    metrics = metric_set(rmse)
  )
```

### Model evaluation
```{r}
rf_tune2%>%
  autoplot()
rf_tune2%>%
  show_best()
rf_best2=rf_tune2%>%
  select_best()
rf_final2=rf_wf2%>%
  finalize_workflow(rf_best2)
rf_fit2=rf_final2%>%
  fit(hosp_data)
rf_predict2=rf_fit2%>%
  predict(hosp_data)
hosp_data%>%
  select(Hospitalizations)%>%
  bind_cols(rf_predict2)%>%
  ggplot(aes(Hospitalizations,.pred))+
  geom_point()+
  geom_abline(intercept = 0,slope=1)+
  geom_smooth(se=FALSE)+
  labs(y="Predicted Hospitalizations",title="Predicted Hospitalizations VS Hospitalizations by random forest model")
ggsave("Predicted Hospitalizations VS Hospitalizations by random forest model.png",path=images_stats_path,width=3000,height=2000,units="px")
```
## Overall, random forest model has the lowesr rmse among those selected models when training data is fitted; Also, the random forest model outperforms other models in predicting outliers.

## Fast and Frugal Trees
## Organize four models performance and choose the best one
## fit testing data
